---
title: "Figure 3, Pre-treatment Comparisons"
author: "S. Grant"
date: "2026-09-03"
output: html_document
---

Format Data
```{r}
rna <- read.csv("~//Documents//COH Breast Cohort//updated_rnaseq_FINAL.csv")
genes <- rna$X
rna <- rna[,-1]
rna <- data.frame(apply(rna, 2, as.numeric))
rna <- log2(rna + 1)
rownames(rna) <- genes
colnames(rna) <- gsub("X.", "/", colnames(rna))

clin <- before2_times
clin$ID <- toupper(clin$ID)
clin <- clin[,c(4,2,3,5:ncol(clin))]
colnames(rna) <- gsub("[.]", "-", colnames(rna))
missing_pretreatment <- setdiff(colnames(rna), clin$ID)
clin <- subset(clin, ID %in% colnames(rna))
```

Establish treatment naive, endocrine treatment only, and chemotherapy treated patient cohorts

```{r}
comb_tests$`specimenID/accessionID` <- toupper(comb_tests$`specimenID/accessionID`)
notreatment <- subset(comb_tests, `specimenID/accessionID` %in% missing_pretreatment)
notreatment <- unique(notreatment)
notreatment$days_since_first_dx <- as.numeric(notreatment$days_since_first_dx)
plot(density(notreatment$days_since_first_dx))
```

```{r}
notreatment2 <- subset(notreatment, days_since_first_dx <= 180)
table(subset(all_data2, UUID %in% notreatment2$UUID)$Stage)
notreatment2 <- subset(notreatment2, UUID %in% subset(all_data2, Stage %in% c("Stage I", "Stage II"))$UUID)
```

```{r}
clin[,4:ncol(clin)] <- apply(clin[,4:ncol(clin)], 2, as.numeric)
### endocrine therapy only
clin$ET_Sum <- as.numeric(rowSums(clin[,grep("Hormone", colnames(clin))]))

clin$Chemo_Sum <- as.numeric(rowSums(clin[,grep("Chemo", colnames(clin))]))

other_cols <- setdiff(colnames(clin), c(colnames(clin)[grep("Hormone", colnames(clin))], "ID", "Stage", "Subtype", "ET_Sum", "Chemo_Sum", "Other_Sum"))
clin$Other_Sum <- as.numeric(rowSums(clin[,other_cols]))
endocrine_only <- subset(clin, ET_Sum > 30 & Other_Sum == 0)

chemo <- subset(clin, Chemo_Sum >= 30)

cdk <- subset(clin, `Duration of CDK Inhibitor` >= 30)

her2 <- subset(clin, `Duration of HER2/ERBB Inhibitor` >= 30)

all_other <- subset(clin, Other_Sum >= 30)
```

Get list of all markers associated with treatment condition from LASSO model
- only running each analysis for subtypes with substantial sample sizes for comparison
- removing non-coding genes from gene expression analyses

RNA-seq
```{r}
all_genes_list <- data.frame()
```


```{r}
hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% endocrine_only$ID] <- "Hormone"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Hormone Therapy"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("VN1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPL", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPS", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  hm_list[[i]] <- hm
  i <- i + 1
}
```


```{r}
cdk_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% cdk$ID] <- "CDK_Inhibitor"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "CDK Inhibitor"))
                          
                          
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("KB.", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPL", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPS", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC1", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5", "BNIP3P29"))
  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  cdk_hm_list[[i]] <- hm
  i <- i + 1
    
}
```

```{r}
her2_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2+", "HR-/HER2+")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% her2$ID] <- "HER2_Inhibitor"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "HER2 Inhibitor"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("KB.", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP6", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  her2_hm_list[[i]] <- hm
  i <- i + 1
    
}
```

```{r}
am_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID] <- "Antimetabolite"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Antimetabolite"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPS", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, c(selected_variables, "Group")]
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  am_hm_list[[i]] <- hm
  i <- i + 1
    
}
```

```{r}
pb_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Platinum Based)` >= 30)$ID] <- "Platinum-Based"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Platinum Based"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("KB.", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, c(selected_variables, "Group")]
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  pb_hm_list[[i]] <- hm
  i <- i + 1
    
}
```


```{r}
antra_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID] <- "Antracycline"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Antracycline"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("LINC", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("FAM", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, c(selected_variables, "Group")]
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  antra_hm_list[[i]] <- hm
  i <- i + 1
    
}
```


```{r}
aa_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID] <- "Alkylating Agent"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Alkylating Agent"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("LINC", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("FAM", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, c(selected_variables, "Group")]
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  aa_hm_list[[i]] <- hm
  i <- i + 1
    
}
```


```{r}
taxane_hm_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  
  data <- data.frame(t(rna))
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID] <- "Taxane"
  
  data <- na.omit(data)
  all_data2$ID <- toupper(all_data2$ID)
  clin2 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin2$ID,]
  
  data$Subtype <- clin2$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  data$Group <- factor(data$Group)
  data$Group <- relevel(data$Group, ref = "Naive")
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  all_genes_list <- rbind(all_genes_list, data.frame("Genes" = non_zero_df$Name,
                                                     "Val" = non_zero_df$Val,
                                                     "Subtype" = subtype,
                                                     "Drug" = "Taxane"))

  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP11", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP13", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP5", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RF0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNV", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTD", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RN7S", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("SNOR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("MIR", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNA", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RNU", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP4", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AP0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC1", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("AC0", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("LINC", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("FAM", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("KB.", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPS", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RPL", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("RP3", selected_variables)])
  selected_variables <- setdiff(selected_variables, selected_variables[grep("CTB", selected_variables)])
  selected_variables <- setdiff(selected_variables, c("AP001052.1", "AC123900.2", "AF254983.5"))
  
  data_filt <- data[, c(selected_variables, "Group")]
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
  column_ha = HeatmapAnnotation(Subtype = data_filt$Subtype,
                                col = list(Subtype = c("HR+/HER2-" = p1_five[1],
                                                       "HR+/HER2+" = p1_five[2],
                                                       "HR-/HER2+" = p1_five[3],
                                                       "HR-/HER2-" = p1_five[4],
                                                       "Not Reported" = "white")))
  data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))] <- scale(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))])
  hm <- Heatmap(t(data_filt[,-c((ncol(data_filt)-2):ncol(data_filt))]), name = "expression",  
          na_col = "white",  top_annotation = column_ha, show_heatmap_legend = T, 
          show_column_names = FALSE, column_split = data_filt$Group,
          cluster_column_slices = T, show_row_names = T, column_title_rot = 45)
  taxane_hm_list[[i]] <- hm
  i <- i + 1
    
}
```

```{r}
summary <- read.csv("~//Downloads//protein_coding_gene_categories_with_summaries.csv")
colnames(summary)[1] <- "Genes"
all_genes <- merge(all_genes_list, summary, by = "Genes")
```

```{r}
all_genes2 <- subset(all_genes, Val > 0)
```

Plot up and downregulated genes (Figure 3a-b)
```{r}
library(ggsankey)
all_genes2 <- unique(all_genes2)
df <- all_genes2 %>%
  make_long(Drug, Genes)

## add color
df$node <- factor(df$node, levels = c(unique(all_genes2$Drug), unique(all_genes2$Category),
                                      unique(all_genes2$Genes)))


colors <- c(pal <- pal_observable("observable10", alpha = 0.9)(8),
            rep("white", length(unique(all_genes2$Genes))))
```

```{r}
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               label = node,
               fill = node)) +
  
  geom_sankey() +
  geom_sankey_label(size = 2, color = 1, fill = "white") +
  theme_sankey(base_size = 16) +
  scale_fill_manual(values = colors) +
  theme(legend.position = "none")
```

```{r}
group <- c()
for (i in 1:nrow(df)){
  if (df$x[i] == 'Genes'){group <- c(group, df$next_node[i])}
  if (df$x[i] == 'Drug'){group <- c(group, as.character(df$node[i]))}
}
df$Group <- group

ls_order <- c()
for (g in unique(all_genes$Category)){
  ls_order <- c(ls_order, unique(subset(all_genes, Category == g)$Genes))
}

df$next_node <- factor(df$next_node, levels = c(ls_order, unique(df$Group)))
df$node <- factor(df$node, levels = c(ls_order, unique(df$Group)))

mycolors <- pal_observable("observable10", alpha = 0.9)(8)

width <- .1
p <- ggplot(df, aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = Group, label = node)) +
  geom_sankey(flow.alpha = 1, node.color = "black", show.legend = FALSE, width = 0.1, linewidth = 0.1) +
  theme_void() +
  theme(
    plot.margin = unit(rep(5.5, 4), "pt")
  ) +
  scale_fill_manual(values = mycolors)


# Get the data from the flows layer
dat <- layer_data(last_plot(), 1) |>
  filter(x ==  2 - width / 2) |>
  distinct(fill, flow_end_ymax, .keep_all = TRUE)

# Get the data from the boxes layer
dat1 <- layer_data(last_plot(), 2) |>
  filter(xmin == 2 - width / 2)
p +
  geom_rect(data = dat, aes(
    xmin = x, xmax = x + width,
    ymin = flow_end_ymin, ymax = flow_end_ymax,
    fill = label
  ), inherit.aes = FALSE) +
  # Draw the outline
  geom_rect(data = dat1, aes(
    xmin = xmin, xmax = xmax,
    ymin = ymin, ymax = ymax
  ), inherit.aes = FALSE, fill = NA, color = NA, linewidth = 0.2) +
  #geom_sankey_text(data = subset(df_phenotypes, x == "Phenotype"), size = 2, color = "black") +
  #geom_sankey_text(data = subset(df_phenotypes, x == "Genes"), size = 2, color = "black") +
  #geom_sankey_label(size = 1, color = "black", fill = "white") +
    geom_sankey_text(
    aes(
      # Align labels conditional on position
      hjust = dplyr::case_when(
        x == "Phenotype" ~ 1,
        x == "Genes" ~ 0
      )
    ),
    size = 3,
    color = "black"
  ) +
  guides(fill = "none")

```

```{r}
all_genes2 <- subset(all_genes, Val < 0)
```

```{r}
library(ggsankey)
all_genes2 <- unique(all_genes2)
df <- all_genes2 %>%
  make_long(Drug, Genes)
```

```{r}
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               label = node,
               fill = node)) +
  
  geom_sankey() +
  geom_sankey_label(size = 2, color = 1, fill = "white") +
  theme_sankey(base_size = 16) +
  scale_fill_observable()+
  #scale_fill_manual(values = colors) +
  theme(legend.position = "none")
```
```{r}
group <- c()
for (i in 1:nrow(df)){
  if (df$x[i] == 'Genes'){group <- c(group, df$next_node[i])}
  if (df$x[i] == 'Drug'){group <- c(group, df$node[i])}
}
df$Group <- group

ls_order <- c()
for (g in unique(all_genes$Category)){
  ls_order <- c(ls_order, intersect(unique(subset(all_genes, Category == g)$Genes), df$node))
}

df$next_node <- factor(df$next_node, levels = c(ls_order, unique(df$Group)))
df$node <- factor(df$node, levels = c(ls_order, unique(df$Group)))

mycolors <- pal_observable("observable10", alpha = 0.9)(8)

width <- .1
p <- ggplot(df, aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = Group, label = node)) +
  geom_sankey(flow.alpha = 1, node.color = "black", show.legend = FALSE, width = 0.1, linewidth = 0.1) +
  theme_void() +
  theme(
    plot.margin = unit(rep(5.5, 4), "pt")
  ) +
  scale_fill_manual(values = mycolors)


# Get the data from the flows layer
dat <- layer_data(last_plot(), 1) |>
  filter(x ==  2 - width / 2) |>
  distinct(fill, flow_end_ymax, .keep_all = TRUE)

# Get the data from the boxes layer
dat1 <- layer_data(last_plot(), 2) |>
  filter(xmin == 2 - width / 2)
p +
  geom_rect(data = dat, aes(
    xmin = x, xmax = x + width,
    ymin = flow_end_ymin, ymax = flow_end_ymax,
    fill = label
  ), inherit.aes = FALSE) +
  # Draw the outline
  geom_rect(data = dat1, aes(
    xmin = xmin, xmax = xmax,
    ymin = ymin, ymax = ymax
  ), inherit.aes = FALSE, fill = NA, color = NA, linewidth = 0.2) +
  #geom_sankey_text(data = subset(df_phenotypes, x == "Phenotype"), size = 2, color = "black") +
  #geom_sankey_text(data = subset(df_phenotypes, x == "Genes"), size = 2, color = "black") +
  #geom_sankey_label(size = 1, color = "black", fill = "white") +
    geom_sankey_text(
    aes(
      # Align labels conditional on position
      hjust = dplyr::case_when(
        x == "Phenotype" ~ 1,
        x == "Genes" ~ 0
      )
    ),
    size = 3,
    color = "black"
  ) +
  guides(fill = "none")

```

```{r}
markers <- rbind(markers, data.frame("Data" = "RNA", "Marker" = all_genes_list$Genes, "Drug" = all_genes_list$Drug))
```

Ecotyper State Abundances

```{r}
annotations <- read.csv("state_annotations.csv")
annotations <- annotations[-1,]

celltypes <- c("B.cells", "CD4.T.cells", "CD8.T.cells", "Dendritic.cells", "Endothelial.cells", "Epithelial.cells",
               "Fibroblasts", "Mast.cells", "Monocytes.and.Macrophages", "PMNs", "NK.cells", "PCs")
celltypes_labels <- unique(annotations$Cell.type)


sa_df <- data.frame()
for (i in 1:length(celltypes)){
  
  celltype <- celltypes[i]
  celltype_label <- celltypes_labels[i]
  
  state_abundance <- read.delim(paste0("~//ecotyper_rnaseq_FINAL//", 
                                       celltype, "//state_abundances.txt"), sep = "\t")
  
  state_abundance$Cell.state <- row.names(state_abundance)
  if (celltype == "PCs"){celltype <- "Plasma cells"}
  anno_sub <- subset(annotations, Cell.type == gsub("[.]", " ",celltype))
  state_abundance <- merge(state_abundance, anno_sub[,c("Cell.type", "Cell.state", "Cell.state.annotation")], by = "Cell.state")
  
  sa_df <- rbind(sa_df, state_abundance)
}
```

```{r}
clin2 <- all_data2
clin2$ID <- gsub("[^[:alnum:] ]", "", clin2$ID)

row.names(sa_df) <- paste0(sa_df$Cell.type, " ", sa_df$Cell.state.annotation)
sa_df <- sa_df[,-c(1,1734:1735)]
sa_df <- data.frame(t(sa_df))
```

```{r}
plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", endocrine_only$ID)] <- "Hormone"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:23, 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Hormone Therapy"))
  
  plot_list[[i]] <- plot
  i <- i + 1
    
}

```


```{r}
cdk_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", cdk$ID)] <- "CDK Inhibitor"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "CDK Inhibitor"))
  cdk_plot_list[[i]] <- plot
  i <- i + 1
    
}

```


```{r}
her2_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2+", "HR-/HER2+")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", her2$ID)] <- "HER2 Inhibitor"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "HER2 Inhibitor"))
  
  her2_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
am_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID)] <- "Antimetabolite"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Antimetabolite"))
  
  am_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
aa_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID)] <- "Alkylating Agent"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Alkylating Agent"))
  
  aa_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
pb_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Platinum Based)` >= 30)$ID)] <- "Platinum-Based"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Platinum-Based"))
  
  pb_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
antra_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID)] <- "Antracycline"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Antracycline"))
  
  antra_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
taxane_plot_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- sa_df
  data$Group <- NA
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
  data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID)] <- "Taxane"
  
  data <- na.omit(data)

  clin3 <- subset(clin2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-3), 
                             names_to = "State", 
                             values_to = "Abundance")
  
  plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~State, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "Ecotyper", "Marker" = unique(data_filt$State), "Drug" = "Taxane"))
  
  taxane_plot_list[[i]] <- plot
  i <- i + 1
    
}
```

ssGSEA pathway scores

```{r}
library(ggplot2)
library(limma)
library(GSVA)
library(msigdbr)
library(dplyr)
library(GSEABase)
library(sva)
```


```{r}
rna <- read.csv("rnaseq.csv")
genes <- rna$X
rna <- rna[,-1]
rna <- data.frame(apply(rna, 2, as.numeric))
rna <- log2(rna + 0.0001)
rownames(rna) <- genes
colnames(rna) <- gsub("X.", "/", colnames(rna))


# calculate ssGSEA scores for Hallmark and KEGG

hallmark_gene_sets <- msigdbr(
  species = "Homo sapiens", # Can change this to what species you need
  collection = "H" # Only hallmark gene sets
)

hallmarks_list <- split(
  hallmark_gene_sets$gene_symbol,
  hallmark_gene_sets$gs_name
)

expression_matrix <- as.matrix(rna)

gsvapar <- ssgseaParam(exprData = expression_matrix,
                                    geneSets = hallmarks_list,
                                    minSize = 1, maxSize = Inf,
                                    alpha = 0.25, normalize = T)

gsea_results <- gsva(param = gsvapar)

hallmark_scores <- as.data.frame(gsea_results)

c2_gene_sets <- msigdbr(
  species = "Homo sapiens",
  collection = "C2"
)

c2_list <- split(
  c2_gene_sets$gene_symbol,
  c2_gene_sets$gs_name
)

gsvapar <- ssgseaParam(exprData = expression_matrix,
                       geneSets = c2_list,
                       minSize = 1, maxSize = Inf,
                       alpha = 0.25, normalize = T)

gsea_results <- gsva(param = gsvapar)

c2_scores <- as.data.frame(gsea_results)

c6_gene_sets <- msigdbr(
  species = "Homo sapiens", # Can change this to what species you need
  collection = "C6"
)

c6_list <- split(
  c6_gene_sets$gene_symbol,
  c6_gene_sets$gs_name
)

gsvapar <- ssgseaParam(exprData = expression_matrix,
                       geneSets = c6_list,
                       minSize = 1, maxSize = Inf,
                       alpha = 0.25, normalize = T)

gsea_results <- gsva(param = gsvapar)

c6_scores <- as.data.frame(gsea_results)

hallmark_scores2 <- as.data.frame(t(hallmark_scores))
c2_scores2 <- as.data.frame(t(c2_scores))
c6_scores2 <- as.data.frame(t(c6_scores))

hallmark_scores2$ID <- rownames(hallmark_scores2)
c2_scores2$ID <- rownames(c2_scores2)
c6_scores2$ID <- rownames(c6_scores2)
```


```{r}
df <- merge(hallmark_scores2, c2_scores2, by = "ID")
df <- merge(df, c6_scores2, by = "ID")

rownames(df) <- df$ID
df <- df[,-1]
pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% endocrine_only$ID] <- "Hormone"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])
  
  if (length(selected_variables) > 20){
    print(selected_variables)
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Hormone Therapy"))
  
  pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
cdk_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% cdk$ID] <- "CDK Inhibitor"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])
  
  if (length(selected_variables) > 20){
    print(selected_variables)
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "CDK Inhibitor"))
  
  cdk_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
aa_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID] <- "Alkylating Agent"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  if (length(selected_variables) > 20){
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Alkylating Agent"))
  
  aa_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
antra_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID] <- "Antracycline"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  if (length(selected_variables) > 20){
    print(selected_variables)
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Antracycline"))
  
  antra_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
am_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-","HR-/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID] <- "Antimetabolite"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  print(selected_variables)
  if (length(selected_variables) > 20){
    
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Antimetabolite"))
  
  am_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
pb_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2+", "HR-/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Platinum Based)` >= 30)$ID] <- "Platinum-Based"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  print(selected_variables)
  if (length(selected_variables) > 20){
    
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Platinum-Based"))
  
  pb_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
taxane_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2-")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID] <- "Taxane"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])

  
  if (length(selected_variables) > 20){
    print(selected_variables)
    non_zero_df$abs <- abs(non_zero_df$Val)
    non_zero_df <- subset(non_zero_df, Name != "(Intercept)")
    non_zero_df <- non_zero_df[order(-non_zero_df$abs),]
    selected_variables <- non_zero_df$Name[1:20]
  }

  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  
  data_filt$Group <- factor(data_filt$Group, levels = c("Taxane", "Naive"))
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y") +
    theme(strip.text = element_text(size = 5))
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "Taxane"))
  
  taxane_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
her2_pathway_list = list()
i <- 1
for (subtype in c("HR+/HER2+", "HR-/HER2+")){
  data <- df
  data$Group <- NA
  data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[rownames(data) %in% her2$ID] <- "HER2 Inhibitor"
  
  data <- na.omit(data)

  clin3 <- subset(all_data2, ID %in% row.names(data))
  data <- data[clin3$ID,]
  
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    data$Group, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  
  selected_variables <- setdiff(selected_variables, selected_variables[grep("Unknown", selected_variables)])
  print(selected_variables)
  
  data_filt <- data[, selected_variables]
  data_filt$Group <- data$Group
  data_filt$Subtype <- subtype
  data_filt <- na.omit(data_filt)

  data_filt <- data_filt %>% tidyr::pivot_longer( 
                             cols = 1:(ncol(data_filt)-2), 
                             names_to = "Pathway", 
                             values_to = "ssGSEA_Score")
  
  plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
    geom_boxplot(position = "dodge2", aes(fill = Group)) + 
    theme_classic() + 
    ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
    facet_wrap(~Pathway, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "GSEA", "Marker" = unique(data_filt$Pathway), "Drug" = "HER2 Inhibitor"))
  
  her2_pathway_list[[i]] <- plot
  i <- i + 1
    
}
```

```{r}
pal <- pal_observable("observable10", alpha = 0.9)(10)
```

Manucript plots for hormone comparison (Figure 3c)
```{r}
df <- merge(hallmark_scores2, c2_scores2, by = "ID")
df <- merge(df, c6_scores2, by = "ID")

rownames(df) <- df$ID
df <- df[,-1]

subtype <- "HR+/HER2-"
data <- df
data$Group <- NA
data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
data$Group[rownames(data) %in% endocrine_only$ID] <- "Hormone"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("KEGG_MEDICUS_REFERENCE_IL1_IL1R_JNK_SIGNALING_PATHWAY",
                        "MARTIN_NFKB_TARGETS_DN", 
                        "TURJANSKI_MAPK11_TARGETS", "TURJANSKI_MAPK8_AND_MAPK9_TARGETS")
  

data_filt <- data[, selected_variables]
data_filt$Group <- data$Group
data_filt$Subtype <- subtype
data_filt <- na.omit(data_filt)

data_filt <- data_filt %>% tidyr::pivot_longer( 
                           cols = 1:(ncol(data_filt)-2), 
                           names_to = "Pathway", 
                           values_to = "ssGSEA_Score")
  
  
  
hormone_pathways_plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
  geom_violin(width=1, position = position_dodge(1), trim = FALSE, color = NA,
              draw_quantiles = c(0.5), quantile.color = "white", quantile.linetype = 0) +
  theme_classic() + 
  ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
  scale_fill_manual(values = c("Hormone" = pal[3], "Naive" = pal[1])) +
  facet_wrap(~Pathway, scale = "free_y", nrow = 2, ncol = 2) +
  theme(strip.text = element_text(size = 6))
hormone_pathways_plot
```

```{r}

subtype = "HR+/HER2-"
data <- sa_df
data$Group <- NA
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", endocrine_only$ID)] <- "Hormone"

data <- na.omit(data)

clin3 <- subset(clin2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("Dendritic.cells.Myeloid.cDC1..S01.",
                        "Dendritic.cells.Myeloid.cDC2.B..Inflammatory...S02.",
                        "Dendritic.cells.Mature..normal.enriched...S05.",
                        "Dendritic.cells.Migratory.activated..S07.",
                        "Mast.cells.Normal.enriched..S02.",
                        "Mast.cells.Activated..S06.")

  
data_filt <- data[, selected_variables]
data_filt$Group <- data$Group
data_filt$Subtype <- subtype
data_filt <- na.omit(data_filt)

data_filt <- data_filt %>% tidyr::pivot_longer( 
                           cols = 1:(ncol(data_filt)-2), 
                           names_to = "State", 
                           values_to = "Abundance")
  
hormone_state_plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
  geom_violin(width=1, position = position_dodge(1), trim = FALSE, color = NA,
              draw_quantiles = c(0.5), quantile.color = "white", quantile.linetype = 0) +
  theme_classic() + 
  ggpubr::stat_compare_means(hide.ns = T, vjust = 1.1) +
  scale_fill_manual(values = c("Hormone" = pal[3], "Naive" = pal[1])) +
  facet_wrap(~State, scale = "free_y", nrow = 2, ncol = 3) +
  theme(strip.text = element_text(size = 6))
hormone_state_plot
```

```{r}
ggpubr::ggarrange(hormone_pathways_plot, hormone_state_plot, ncol = 2, align = "hv", common.legend = T,
                  widths = c(0.66, 1))
```

Plot SMARCB1 for Antimetabolite, alkylating Agent, Antracycline, and Taxane (Supplementary Figure)

```{r}
df <- merge(hallmark_scores2, c2_scores2, by = "ID")
df <- merge(df, c6_scores2, by = "ID")

rownames(df) <- df$ID
df <- df[,-1]

subtype <- "HR+/HER2-"
data <- df
data$Group <- NA
data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID] <- "Taxane"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("WP_TUMOR_SUPPRESSOR_ACTIVITY_OF_SMARCB1")
  
data_filt <- data[, c(selected_variables, "Group")]
data_filt$Subtype <- subtype
data_filt <- na.omit(data_filt)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID] <- "Alkylating Agent"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("WP_TUMOR_SUPPRESSOR_ACTIVITY_OF_SMARCB1")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID] <- "Antimetabolite"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("WP_TUMOR_SUPPRESSOR_ACTIVITY_OF_SMARCB1")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID] <- "Antracycline"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("WP_TUMOR_SUPPRESSOR_ACTIVITY_OF_SMARCB1")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)


data_filt <- data_filt %>% tidyr::pivot_longer( 
                           cols = 1:(ncol(data_filt)-2), 
                           names_to = "Pathway", 
                           values_to = "ssGSEA_Score")
  
data_filt$Group <- factor(data_filt$Group)
data_filt$Group <- relevel(data_filt$Group, ref = "Naive")
  
smarcb1_pathways_plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
  geom_violin(width=1, position = position_dodge(1), trim = FALSE, color = NA,
              draw_quantiles = c(0.5), quantile.color = "white", quantile.linetype = 0) +
  theme_classic() + 
  ggpubr::stat_compare_means(comparisons = list(c("Naive", "Alkylating Agent"), c("Naive", "Antimetabolite"),
                                                c("Naive", "Antracycline"), c("Naive", "Taxane")),
                             hide.ns = T) +
  scale_fill_manual(values = c("Naive" = pal[1], "Alkylating Agent" = pal[2], "Antimetabolite" = pal[3],
                               "Antracycline" = pal[4], "Taxane" = pal[5])) +
  facet_wrap(~Pathway, scale = "free_y") +
  theme(strip.text = element_text(size = 6))
smarcb1_pathways_plot
```


Manuscript plots Chemotherapy treated vs treatment-naive (Figure 3e-f)


```{r}
df <- merge(hallmark_scores2, c2_scores2, by = "ID")
df <- merge(df, c6_scores2, by = "ID")

rownames(df) <- df$ID
df <- df[,-1]

subtype <- "HR+/HER2-"
data <- df
data$Group <- NA
data$Group[rownames(data) %in% notreatment2$`specimenID/accessionID`] <- "Naive"
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID] <- "Taxane"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("KEGG_MEDICUS_REFERENCE_TNF_RIPK1_3_SIGNALING_PATHWAY",
                        "REACTOME_DEFECTIVE_RIPK1_MEDIATED_REGULATED_NECROSIS",
                        "REACTOME_TLR3_MEDIATED_TICAM1_DEPENDENT_PROGRAMMED_CELL_DEATH")
  

data_filt <- data[, c(selected_variables, "Group")]
data_filt$Subtype <- subtype
data_filt <- na.omit(data_filt)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID] <- "Alkylating Agent"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("KEGG_MEDICUS_REFERENCE_TNF_RIPK1_3_SIGNALING_PATHWAY",
                        "REACTOME_DEFECTIVE_RIPK1_MEDIATED_REGULATED_NECROSIS",
                        "REACTOME_TLR3_MEDIATED_TICAM1_DEPENDENT_PROGRAMMED_CELL_DEATH")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID] <- "Antimetabolite"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("KEGG_MEDICUS_REFERENCE_TNF_RIPK1_3_SIGNALING_PATHWAY",
                        "REACTOME_DEFECTIVE_RIPK1_MEDIATED_REGULATED_NECROSIS",
                        "REACTOME_TLR3_MEDIATED_TICAM1_DEPENDENT_PROGRAMMED_CELL_DEATH")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data <- df
data$Group <- NA
data$Group[rownames(data) %in% subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID] <- "Antracycline"

data <- na.omit(data)

clin3 <- subset(all_data2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("KEGG_MEDICUS_REFERENCE_TNF_RIPK1_3_SIGNALING_PATHWAY",
                        "REACTOME_DEFECTIVE_RIPK1_MEDIATED_REGULATED_NECROSIS",
                        "REACTOME_TLR3_MEDIATED_TICAM1_DEPENDENT_PROGRAMMED_CELL_DEATH")

data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)


data_filt <- data_filt %>% tidyr::pivot_longer( 
                           cols = 1:(ncol(data_filt)-2), 
                           names_to = "Pathway", 
                           values_to = "ssGSEA_Score")
  
data_filt$Group <- factor(data_filt$Group)
data_filt$Group <- relevel(data_filt$Group, ref = "Naive")
  
chemo_pathways_plot <- ggplot(data_filt, aes(x = Group, y = ssGSEA_Score, fill = Group)) +
  geom_violin(width=1, position = position_dodge(1), trim = FALSE, color = NA,
              draw_quantiles = c(0.5), quantile.color = "white", quantile.linetype = 0) +
  theme_classic() + 
  ggpubr::stat_compare_means(comparisons = list(c("Naive", "Alkylating Agent"), c("Naive", "Antimetabolite"),
                                                c("Naive", "Antracycline"), c("Naive", "Taxane")),
                             hide.ns = T) +
  scale_fill_manual(values = c("Naive" = pal[1], "Alkylating Agent" = pal[2], "Antimetabolite" = pal[3],
                               "Antracycline" = pal[4], "Taxane" = pal[5])) +
  facet_wrap(~Pathway, scale = "free_y") +
  theme(strip.text = element_text(size = 6))
chemo_pathways_plot
```
```{r}

subtype = "HR+/HER2-"
data <- sa_df
data$Group <- NA
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", notreatment2$`specimenID/accessionID`)] <- "Naive"
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID)] <- "Taxane"

data <- na.omit(data)

clin3 <- subset(clin2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("CD4.T.cells.Naïve.central.memory..S02.",
                        "CD8.T.cells.Naïve.central.memory..S01.")

  
data_filt <- data[, c(selected_variables, "Group")]
data_filt$Subtype <- subtype
data_filt <- na.omit(data_filt)

data <- sa_df
data$Group <- NA
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID)] <- "Alkylating Agent"

data <- na.omit(data)

clin3 <- subset(clin2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("CD4.T.cells.Naïve.central.memory..S02.",
                        "CD8.T.cells.Naïve.central.memory..S01.")

  
data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data <- sa_df
data$Group <- NA
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID)] <- "Antimetabolite"

data <- na.omit(data)

clin3 <- subset(clin2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("CD4.T.cells.Naïve.central.memory..S02.",
                        "CD8.T.cells.Naïve.central.memory..S01.")

  
data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)


data <- sa_df
data$Group <- NA
data$Group[rownames(data) %in% gsub("[^[:alnum:] ]", "", subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID)] <- "Antracycline"

data <- na.omit(data)

clin3 <- subset(clin2, ID %in% row.names(data))
data <- data[clin3$ID,]

data$Subtype <- clin3$Subtype
data <- subset(data, Subtype == subtype)[,-ncol(data)]

selected_variables <- c("CD4.T.cells.Naïve.central.memory..S02.",
                        "CD8.T.cells.Naïve.central.memory..S01.")

  
data_filt2 <- data[, c(selected_variables, "Group")]
data_filt2$Subtype <- subtype
data_filt2 <- na.omit(data_filt2)
data_filt <- rbind(data_filt, data_filt2)

data_filt <- data_filt %>% tidyr::pivot_longer( 
                           cols = 1:(ncol(data_filt)-2), 
                           names_to = "State", 
                           values_to = "Abundance")
  
data_filt$Group <- factor(data_filt$Group)
data_filt$Group <- relevel(data_filt$Group, ref = "Naive")
  
chemo_state_plot <- ggplot(data_filt, aes(x = Group, y = Abundance, fill = Group)) +
  geom_violin(width=1, position = position_dodge(1), trim = FALSE, color = NA,
              draw_quantiles = c(0.5), quantile.color = "white", quantile.linetype = 0) +
  theme_classic() + 
  ggpubr::stat_compare_means(comparisons = list(c("Naive", "Alkylating Agent"), c("Naive", "Antimetabolite"),
                                                c("Naive", "Antracycline"), c("Naive", "Taxane")),
                             hide.ns = T) +
  scale_fill_manual(values = c("Naive" = pal[1], "Alkylating Agent" = pal[2], "Antimetabolite" = pal[3],
                               "Antracycline" = pal[4], "Taxane" = pal[5])) +
  facet_wrap(~State, scale = "free_y") +
  theme(strip.text = element_text(size = 6))
chemo_state_plot
```


```{r}
chemo_plots <- ggpubr::ggarrange(chemo_pathways_plot, chemo_state_plot, ncol = 2, nrow = 1,
                  align = "hv", common.legend = T,
                  widths = c(1, 0.75))
hormone_plots <- ggpubr::ggarrange(hormone_pathways_plot, hormone_state_plot, ncol = 2, align = "hv", common.legend = T,
                  widths = c(0.66, 1))

ggpubr::ggarrange(hormone_plots, chemo_plots, nrow = 2, align = "hv", heights = c(1, 0.7))
```



Genomic Data

Organize datasets (convert to gene-level mutation and CNA features for LASSO model)
```{r}
mutations <- read.csv("coh_breast_cohort_snv_updated.csv")
gain <- read.csv("coh_breast_cohort_cna_gain_v2.csv")
loss <- read.csv("coh_breast_cohort_cna_loss_v2.csv")

ids <- read.csv("uuid_studyid_match_updated.csv")
ids_original <- rbind(read_excel("HB-1984_CR_PCDM_specimenID.xlsx", sheet = " CR_Tests "),
                      read_excel("HB-1984_CR_PCDM_specimenID.xlsx", sheet = "PCDM_Test"))
colnames(ids_original)[3] <- colnames(ids)[1]
ids_original$accessionNumber <- toupper(ids_original$accessionNumber)
ids_original <- subset(ids_original, accessionNumber %in% ids$accessionNumber)
ids2 <- merge(ids, ids_original[,c(1,3)], by = "accessionNumber")

ids2 <- ids2 %>% 
  arrange(match(ids2$StudyID, ids$StudyID))

colnames(mutations) <- c("StudyID", paste0("SNV_", colnames(mutations)[-1]))
colnames(gain) <- c(paste0("GAIN_", colnames(gain)[-ncol(gain)]), "StudyID")
colnames(loss) <- c(paste0("LOSS_", colnames(loss)[-ncol(loss)]), "StudyID")

cna <- merge(gain, loss, by = "StudyID")
mutations <- merge(mutations, cna, by = "StudyID")

mutations <- merge(mutations, ids2, by = "StudyID")
mutations <- mutations[,-c(1,ncol(mutations))]
```

```{r}
clin <- before2_times
clin$ID <- toupper(clin$ID)

clin <- clin[,c(4,2,3,5:ncol(clin))]

colnames(rna) <- gsub("[.]", "-", colnames(rna))

missing_pretreatment <- setdiff(colnames(rna), clin$ID)

clin <- subset(clin, ID %in% mutations$accessionNumber)

clin[,4:ncol(clin)] <- apply(clin[,4:ncol(clin)], 2, as.numeric)
### endocrine therapy only
clin$ET_Sum <- as.numeric(rowSums(clin[,grep("Hormone", colnames(clin))]))

clin$Chemo_Sum <- as.numeric(rowSums(clin[,grep("Chemo", colnames(clin))]))

other_cols <- setdiff(colnames(clin), c(colnames(clin)[grep("Hormone", colnames(clin))], "ID", "Stage", "Subtype", "ET_Sum", "Chemo_Sum", "Other_Sum"))
clin$Other_Sum <- as.numeric(rowSums(clin[,other_cols]))
endocrine_only <- subset(clin, ET_Sum > 30 & Other_Sum == 0)

chemo <- subset(clin, Chemo_Sum >= 30)

cdk <- subset(clin, `Duration of CDK Inhibitor` >= 30)

her2 <- subset(clin, `Duration of HER2/ERBB Inhibitor` >= 30)

all_other <- subset(clin, Other_Sum >= 30)
```

```{r}
mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% endocrine_only$ID] <- "Hormone"
  
  data <- na.omit(unique(data))

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Hormone Therapy"))

  mut_list[[i]] <- plot
  i <- i + 1
}
```



```{r}
cdk_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% cdk$ID] <- "CDK Inhibitor"
  
  data <- na.omit(unique(data))

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "CDK Inhibitor"))

  cdk_mut_list[[i]] <- plot
  i <- i + 1
}
```


```{r}
am_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR+/HER2+")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% subset(clin, `Duration of Chemotherapy (Antimetabolite)` >= 30)$ID] <- "Antimetabolite"
  
  data <- na.omit(unique(data))
  

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  if (length(selected_variables) > 20){
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Antimetabolite"))

  am_mut_list[[i]] <- plot
  i <- i + 1
}
```

```{r}
aa_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-","HR-/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% subset(clin, `Duration of Chemotherapy (Alkylating Agent)` >= 30)$ID] <- "Alylating Agent"
  
  data <- na.omit(unique(data))
  

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  if (length(selected_variables) > 20){
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Alkylating Agent"))

  aa_mut_list[[i]] <- plot
  i <- i + 1
}
```

```{r}
antra_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-",  "HR-/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% subset(clin, `Duration of Chemotherapy (Antracycline)` >= 30)$ID] <- "Antracycline"
  
  data <- na.omit(unique(data))
  

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  if (length(selected_variables) > 20){
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Antracycline"))

  antra_mut_list[[i]] <- plot
  i <- i + 1
}
```



```{r}
pb_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2+", "HR-/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% subset(clin, `Duration of Chemotherapy (Platinum Based)` >= 30)$ID] <- "Platinum Based"
  
  data <- na.omit(unique(data))
  

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  if (length(selected_variables) > 20){
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Platinum-Based"))

  pb_mut_list[[i]] <- plot
  i <- i + 1
}
```


```{r}
taxane_mut_list = list()
i <- 1
for (subtype in c("HR+/HER2-", "HR-/HER2+", "HR-/HER2-")){
  print(subtype)
  data <- mutations
  data$Group <- NA
  data$Group[data$accessionNumber %in% notreatment2$`specimenID/accessionID`] <- "Naive"
  data$Group[data$accessionNumber %in% subset(clin, `Duration of Chemotherapy (Taxane)` >= 30)$ID] <- "Taxane"
  
  data <- na.omit(unique(data))
  

  clin3 <- subset(all_data2, ID %in% data$accessionNumber)
  
  data <- subset(data, accessionNumber %in% clin3$ID)
  row.names(clin3) <- clin3$ID
  clin3 <- clin3[data$accessionNumber,]
  data$Subtype <- clin3$Subtype
  data <- subset(data, Subtype == subtype)[,-ncol(data)]
  
  groups <- data$Group
  
  data <- data[,-c((ncol(data)-1):ncol(data))]
  
  data <- data.frame(apply(data, 2, as.numeric))
  data <- data[, apply(data, 2, var, na.rm=TRUE) != 0]
  # hormone vs nontreated
  set.seed(1)
  fit1 <- cv.glmnet(as.matrix(data[,-ncol(data)]), 
                    groups, family = "binomial", alpha = 1)
  coef_matrix <- as.matrix(coef(fit1, s = "lambda.min"))
  non_zero_vars <- coef_matrix[coef_matrix[, 1] != 0, ]
  non_zero_df <- data.frame("Val" = as.numeric(non_zero_vars), "Name" = names(non_zero_vars))
  non_zero_df <- non_zero_df[order(-non_zero_df$Val),]
  selected_variables <- names(non_zero_vars)[names(non_zero_vars) != "(Intercept)"]
  print(selected_variables)
  data <- data[,selected_variables]
  data$Group <- groups
  
  data <- data %>% tidyr::pivot_longer(
    cols = 1:(ncol(data)-1),
    names_to = "Gene",
    values_to = "Mutations"
  )
  
  data_freq <- data.frame()
  for (gene in unique(data$Gene)){
    g1 <- unique(data$Group)[1]
    g2 <- unique(data$Group)[2]
    freq1 <- sum(subset(data, Group == g1 & Gene == gene)$Mutations) / nrow(subset(data, Group == g1 & Gene == gene))
    freq2 <- sum(subset(data, Group == g2 & Gene == gene)$Mutations) / nrow(subset(data, Group == g2 & Gene == gene))
    row <- data.frame("Gene" = gene, "Frequency" = c(freq1, freq2),
                      "Group" = c(g1, g2))
    data_freq <- rbind(data_freq, row)
  }
  
  data_freq$Group <- factor(data_freq$Group)
  data_freq$Group <- relevel(data_freq$Group, ref = "Naive")
  
  if (length(selected_variables) > 20){
  data_freq <- subset(data_freq, Gene %in% subset(data_freq, Group != "Naive" & Frequency >= 0.05)$Gene)
  }
  
  plot <- ggplot(data_freq, aes(x = Group, fill = Group, y = Frequency)) +
    geom_col(stat = "identity", position = "dodge") +
    theme_classic() +
    facet_wrap(~Gene, scale = "free_y")
  
  markers <- rbind(markers, data.frame("Data" = "DNA", "Marker" = unique(data_freq$Gene), "Drug" = "Taxane"))

  taxane_mut_list[[i]] <- plot
  i <- i + 1
}
```

Save files for mutationmapper input (Figure 3g)

```{r}
vcf <- read.csv("~//Documents//COH Breast Cohort//Files for UCSD//coh_bc_merged_vcf.csv")
colnames(vcf) <- vcf[116,]
vcf <- vcf[-c(1:116),]

vcf <- merge(vcf, ids2, by = "StudyID")

get_genename <- function(x){
  x2 <- strsplit(x, split = "[|]")
  x3 <- x2[[1]][4]
  return(x3)
}
vcf$Gene <- lapply(vcf$INFO, get_genename)

cdk_vcf <- subset(vcf, accessionNumber %in% cdk$ID)
cdk_vcf <- subset(cdk_vcf, Gene %in% c("ESR1"))
```


