---
title: "COH Breast Cohort (Figure 2)"
author: "S. Grant"
date: "2025-08-11"
output: html_document
---

Palettes
```{r}
p1_light <- c("#C2697FFF",  "#E68E54FF", "#FBC559FF","#7FC269FF", "#4151B0FF", "#785596FF")

p1_dark <- c("#9E4058FF", "#D0641EFF","#F9AB0EFF", "#589E40FF",  "#2C3778FF", "#513965FF")

p1_three <- c("#C2697FFF",  "#FBC559FF", "#4151B0FF")

p1_five <- c("#C2697FFF",  "#FBC559FF",  "#7FC269FF", "#4151B0FF",  "#785596FF")
```


```{r}
library(readxl)
library(ggpubr)
library(ggplot2)
library(survival)
library(survminer)
library(ggrepel)
library(dplyr)
library(paletteer)
```
Retrieve testing timepoints
```{r}
cr_tests <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "CR_Tests")
cr_tests2 <- read_excel("~//Documents//COH Breast Cohort//HB-1984_CR_PCDM_specimenID.xlsx", sheet = " CR_Tests ")

colnames(cr_tests) <- cr_tests[1,]
cr_tests <- cr_tests[-1,]

cr_tests <- subset(cr_tests, genetic_test %in% c("ashion", "Guardant", "Tempus"))

pcdm_tests <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "PCDM_Tests")
pcdm_tests2 <- read_excel("~//Documents//COH Breast Cohort//HB-1984_CR_PCDM_specimenID.xlsx", sheet = "PCDM_Test")

colnames(pcdm_tests) <- pcdm_tests[1,]
pcdm_tests <- pcdm_tests[-1,]

pcdm_tests <- subset(pcdm_tests, genetic_test %in% c("ashion", "Guardant", "Tempus"))
```
Retrieve clinical variables and data to calculate overall survival
```{r}
# AGE & OS
pcdm_cohort <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "PCDM_Cohort")
cr_cohort <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "CR_Cohort")

colnames(pcdm_cohort) <- pcdm_cohort[1,]
pcdm_cohort <- pcdm_cohort[-1,]

colnames(cr_cohort) <- cr_cohort[1,]
cr_cohort <- cr_cohort[-1,]

pcdm_cohort <- pcdm_cohort[,c(1,5:6)]
cr_cohort <- cr_cohort[,c(1,4:7)]
cr_cohort <- unique(cr_cohort)

# STAGE
pcdm_stage <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "PCDM_Stage")
cr_stage <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "CR_Stage")

colnames(pcdm_stage) <- pcdm_stage[1,]
pcdm_stage <- pcdm_stage[-1,]

colnames(cr_stage) <- cr_stage[1,]
cr_stage <- cr_stage[-1,]

pcdm_stage <- merge(pcdm_stage, pcdm_cohort, by = "UUID")
pcdm_stage <- unique(pcdm_stage)

pcdm_stage$days_Dx_to_stage <- as.numeric(pcdm_stage$days_Dx_to_stage)
pcdm_stage$days_Dx_to_Death <- as.numeric(pcdm_stage$days_Dx_to_Death)
pcdm_stage$days_Dx_to_LastAppt <- as.numeric(pcdm_stage$days_Dx_to_LastAppt)
pcdm_stage$stage[pcdm_stage$stage == "NULL"] <- "Stage Unknown"
```

Clean up stage - keep most recently updated staging information
```{r}
pcdm_keep <- data.frame()
for (patient in unique(pcdm_stage$UUID)){
  sub <- subset(pcdm_stage, UUID == patient)
  sub2 <- subset(sub, stage != "Stage Unknown" & stage != "No Stage Recommended")
  if (nrow(sub2) > 0){
    keep_row <- sub2[which.max(abs(sub2$days_Dx_to_stage)),]
  }
  if (nrow(sub2) == 0){
    keep_row <- sub[which.max(abs(sub$days_Dx_to_stage)),]
  }
  pcdm_keep <- rbind(pcdm_keep, keep_row)
}

cr_keep_cohort <- subset(cr_cohort, diagnosis_day == 0)
cr_keep <- data.frame()
for (patient in unique(cr_stage$UUID)){
  sub <- subset(cr_stage, UUID == patient)
  sub2 <- subset(sub, Stage != "Stage Unknown")
  sub2$Stage <- sort(sub2$Stage)
  if (nrow(sub2) > 0){
    keep_row <- sub2[nrow(sub2),]
  }
  if (nrow(sub2) == 0){
    keep_row <- sub[nrow(sub),]
  }
  cr_keep <- rbind(cr_keep, keep_row)
}

cr_keep <- merge(cr_keep, cr_keep_cohort, by = "UUID")
```

Clean dataframes - remove unnecessary columns
```{r}
cr_keep <- cr_keep[,c("UUID", "Stage", "ER", "PR", "HER2", "age_diagnosis", "days_Dx_to_Death", "days_Dx_to_LastAppt")]
pcdm_keep <- pcdm_keep[,c("UUID", "age_at_stage", "stage", "erStatus", "prStatus", "her2Status", "days_Dx_to_Death", "days_Dx_to_LastAppt")]

colnames(cr_keep) <- c("UUID", "Stage", "ER", "PR", "HER2", "Age", "OS_Event", "OS_Time")
colnames(pcdm_keep) <- c("UUID", "Age", "Stage", "ER", "PR", "HER2", "OS_Event", "OS_Time")
```

Clean up staging and subtype labels
```{r}
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IIIB"] <- "Stage III"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IIB"] <- "Stage II"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IB"] <- "Stage I"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IA"] <- "Stage I"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IIA"] <- "Stage II"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IIIA"] <- "Stage III"
pcdm_keep$Stage[pcdm_keep$Stage == "Stage IIIC"] <- "Stage III"
pcdm_keep$Stage[pcdm_keep$Stage == "No Stage Recommended"] <- "Not Reported"

cr_keep$Stage[cr_keep$Stage == "Stage 1"] <- "Stage I"
cr_keep$Stage[cr_keep$Stage == "Stage 2"] <- "Stage II"
cr_keep$Stage[cr_keep$Stage == "Stage 3"] <- "Stage III"
cr_keep$Stage[cr_keep$Stage == "Stage 4"] <- "Stage IV"
cr_keep$Stage[cr_keep$Stage == "No Stage Recommended"] <- "Not Reported"

pcdm_keep$ER[pcdm_keep$ER == "NULL" | pcdm_keep$ER == "Not assessed"] <- NA
pcdm_keep$PR[pcdm_keep$PR == "NULL" | pcdm_keep$PR == "Not assessed"] <- NA
pcdm_keep$HER2[pcdm_keep$HER2 == "NULL" | pcdm_keep$HER2 == "Not assessed" | pcdm_keep$HER2 == "Indeterminate/Borderline"] <- NA

pcdm_keep$Subtype <- NA
pcdm_keep$Subtype[pcdm_keep$ER == "Positive" & pcdm_keep$HER2 == "Negative"] <- "HR+/HER2-"
pcdm_keep$Subtype[pcdm_keep$PR == "Positive" & pcdm_keep$HER2 == "Negative"] <- "HR+/HER2-"
pcdm_keep$Subtype[pcdm_keep$ER == "Positive" & pcdm_keep$HER2 == "Positive"] <- "HR+/HER2+"
pcdm_keep$Subtype[pcdm_keep$PR == "Positive" & pcdm_keep$HER2 == "Positive"] <- "HR+/HER2+"
pcdm_keep$Subtype[pcdm_keep$PR == "Negative" & pcdm_keep$ER == "Negative" & pcdm_keep$HER2 == "Positive"] <- "HR-/HER2+"
pcdm_keep$Subtype[pcdm_keep$ER == "Negative" & pcdm_keep$HER2 == "Negative" & pcdm_keep$PR == "Negative"] <- "HR-/HER2-"

cr_keep$ER[cr_keep$ER == "NULL"] <- NA
cr_keep$PR[cr_keep$PR == "NULL"] <- NA
cr_keep$HER2[cr_keep$HER2 == "NULL"] <- NA

cr_keep$Subtype <- NA
cr_keep$Subtype[cr_keep$ER == "Positive" & cr_keep$HER2 == "Negative"] <- "HR+/HER2-"
cr_keep$Subtype[cr_keep$PR == "Positive" & cr_keep$HER2 == "Negative"] <- "HR+/HER2-"
cr_keep$Subtype[cr_keep$ER == "Positive" & cr_keep$HER2 == "Positive"] <- "HR+/HER2+"
cr_keep$Subtype[cr_keep$PR == "Positive" & cr_keep$HER2 == "Positive"] <- "HR+/HER2+"
cr_keep$Subtype[cr_keep$PR == "Negative" & cr_keep$ER == "Negative" & cr_keep$HER2 == "Positive"] <- "HR-/HER2+"
cr_keep$Subtype[cr_keep$ER == "Negative" & cr_keep$HER2 == "Negative" & cr_keep$PR == "Negative"] <- "HR-/HER2-"
```

Calculate overall survival information
```{r}
event <- c()
time <- c()
for (i in 1:nrow(pcdm_keep)){
  if (is.na(pcdm_keep$OS_Event)[i] == TRUE){
    event <- c(event, 0)
    time <- c(time, pcdm_keep$OS_Time[i])
  }
  if (is.na(pcdm_keep$OS_Event)[i] == FALSE){
    event <- c(event, 1)
    time <- c(time, pcdm_keep$OS_Event[i])
  }  
}
pcdm_keep$OS_Event <- unlist(event)
pcdm_keep$OS_Time <- unlist(time)

event <- c()
time <- c()
for (i in 1:nrow(cr_keep)){
  if (cr_keep$OS_Event[i] == "NULL"){
    event <- c(event, 0)
    time <- c(time, cr_keep$OS_Time[i])
  }
  if (cr_keep$OS_Event[i] != "NULL"){
    event <- c(event, 1)
    time <- c(time, cr_keep$OS_Event[i])
  }  
}
cr_keep$OS_Event <- unlist(event)
cr_keep$OS_Time <- unlist(time)
```

Load and clean treatment history data with grouped labels
```{r}
pcdm_med <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "PCDM_Meds")
cr_med <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "CR_Meds")

colnames(pcdm_med) <- pcdm_med[1,]
pcdm_med <- pcdm_med[-1,]

colnames(cr_med) <- cr_med[1,]
cr_med <- cr_med[-1,]

drug_list <- read.csv("MedClass_updated_V3_20250604.csv")
drug_list$Drug_Classification_V4[drug_list$Merged_med_name == "Trastuzumab Deruxtecan"] <- "HER2_ERBB_Inhibitor (ADC)"
drug_list <- drug_list[,-c(1,2)]
colnames(drug_list)[2] <- "med_name"

pcdm_med <- merge(pcdm_med[,1:3], drug_list[,c(2,5)], by = "med_name")
cr_med <- merge(cr_med[,1:3], drug_list[,c(2,5)], by = "med_name")
```


```{r}
cr_tempus <- read_excel("~//Documents//COH Breast Cohort//HB-1984-CR-updated.xlsx", sheet = "Tempus")
cr_tempus <- subset(cr_tempus, significance != "RNA") # remove RNA biomarkers from Tempus assays
cr_guardant <- read_excel("~//Documents//COH Breast Cohort//HB-1984-CR-updated.xlsx", sheet = "Guardant")
cr_ashion <- read_excel("~//Documents//COH Breast Cohort//HB-1984-CR-updated.xlsx", sheet = "GemExtra")

cr_samples <- unique(c(cr_tempus$accessionNumber, cr_guardant$accessionNumber, cr_ashion$specimenNumber))
```

Assess combination of sequencing tests available per patient
```{r}
pcdm_tests <- subset(pcdm_tests2, source %in% c("ashion", "Guardant", "Tempus", "Fulgent", "Invitae"))
colnames(pcdm_tests)[2] <- "genetic_test"
pcdm_tests$genetic_test[pcdm_tests$genetic_test == "ashion"] <- "Ashion"

ful <- c()
inv <- c()
temp <- c()
ash <- c()
guard <- c()
for (u in unique(pcdm_tests$uuid)){
  sub <- subset(pcdm_tests, uuid == u)
  if ("Fulgent" %in% sub$genetic_test){ful <- c(ful, "Yes")}
  else{ful <- c(ful, "No")}
  if ("Invitae" %in% sub$genetic_test){inv <- c(inv, "Yes")}
  else{inv <- c(inv, "No")}
  if ("Tempus" %in% sub$genetic_test){temp <- c(temp, "Yes")}
  else{temp <- c(temp, "No")}
  if ("Ashion" %in% sub$genetic_test){ash <- c(ash, "Yes")}
  else{ash <- c(ash, "No")}
  if ("Guardant" %in% sub$genetic_test){guard <- c(guard, "Yes")}
  else{guard <- c(guard, "No")}
}

tests <- data.frame("UUID" = unique(pcdm_tests$uuid), "Fulgent" = unlist(ful), "Invitae" = unlist(inv), "Tempus" = unlist(temp), "Ashion" = unlist(ash), "Guardant" = unlist(guard))

pcdm_keep <- subset(pcdm_keep, UUID %in% tests$UUID)
pcdm_keep <- merge(pcdm_keep, tests, by = "UUID")
```

Remove patients that do not have somatic sequencing data available
```{r}
colnames(cr_tests2)[2] <- "genetic_test"
cr_tests <- subset(cr_tests2, genetic_test %in% c("Fulgent", "Invitae", "Tempus", "ashion", "Guardant"))
cr_tests$genetic_test[cr_tests$genetic_test == "ashion"] <- "Ashion"

ful <- c()
inv <- c()
temp <- c()
ash <- c()
guard <- c()
for (u in unique(cr_tests$uuid)){
  sub <- subset(cr_tests, uuid == u)
  if ("Fulgent" %in% sub$genetic_test){ful <- c(ful, "Yes")}
  else{ful <- c(ful, "No")}
  if ("Invitae" %in% sub$genetic_test){inv <- c(inv, "Yes")}
  else{inv <- c(inv, "No")}
  if ("Tempus" %in% sub$genetic_test){temp <- c(temp, "Yes")}
  else{temp <- c(temp, "No")}
  if ("Ashion" %in% sub$genetic_test){ash <- c(ash, "Yes")}
  else{ash <- c(ash, "No")}
  if ("Guardant" %in% sub$genetic_test){guard <- c(guard, "Yes")}
  else{guard <- c(guard, "No")}
}

tests <- data.frame("UUID" = unique(cr_tests$uuid), "Fulgent" = unlist(ful), "Invitae" = unlist(inv), "Tempus" = unlist(temp), "Ashion" = unlist(ash), "Guardant" = unlist(guard))

cr_keep <- subset(cr_keep, UUID %in% tests$UUID)
cr_keep <- merge(cr_keep, tests, by = "UUID")
```

```{r}
all_data <- rbind(pcdm_keep, cr_keep)
all_data <- subset(all_data, Tempus == "Yes" | Ashion == "Yes" | Guardant == "Yes")

all_data$Stage[all_data$Stage == "Stage Unknown"] <- "Not Reported"
```

Plot stage (Figure 1a)
```{r}
stage_df <- data.frame(table(all_data$Stage))
colnames(stage_df) <- c("Stage", "Count")
stage_df$Fraction <- (stage_df$Count / sum(stage_df$Count))*100

# Compute the cumulative percentages (top of each rectangle)
stage_df$ymax <- cumsum(stage_df$Fraction)

# Compute the bottom of each rectangle
stage_df$ymin <- c(0, head(stage_df$ymax, n=-1))

# Compute label position
stage_df$labelPosition <- (stage_df$ymax + stage_df$ymin) /2

# Compute a good label
stage_df$label <- paste0(stage_df$Stage, "\n n = ", stage_df$Count)
```

```{r}
# Get the positions
stage_df2 <- stage_df %>% 
  mutate(csum = rev(cumsum(rev(Count))), 
         pos = Count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Count/2, pos))
```


```{r}
stage_plot <- ggplot(stage_df, aes(x = "", y = Count, fill=Stage)) +
  geom_col(width = 1, color = NA, alpha = 0.9) +
  coord_polar(theta = "y") +
  scale_fill_observable()+
  geom_text_repel(data = stage_df2,
                 aes(y = pos, label = label),
                size = 5, nudge_x = 0.85, show.legend = FALSE, fill = NA,
                segment.color = NA)+
  theme_void() +
  theme(legend.position = "none")
stage_plot
```

Plot subtype (Figure 1b)
```{r}
subtype_df <- data.frame(table(all_data$Subtype))
colnames(subtype_df) <- c("Subtype", "Count")
subtype_df$Fraction <- (subtype_df$Count / sum(subtype_df$Count))*100

# Compute the cumulative percentages (top of each rectangle)
subtype_df$ymax <- cumsum(subtype_df$Fraction)

# Compute the bottom of each rectangle
subtype_df$ymin <- c(0, head(subtype_df$ymax, n=-1))

# Compute label position
subtype_df$labelPosition <- (subtype_df$ymax + subtype_df$ymin) /2

# Compute a good label
subtype_df$label <- paste0(subtype_df$Subtype, "\n n = ", subtype_df$Count)
```

```{r}
# Get the positions
subtype_df2 <- subtype_df %>% 
  mutate(csum = rev(cumsum(rev(Count))), 
         pos = Count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Count/2, pos))
```


```{r}
subtype_plot <- ggplot(subtype_df, aes(x = "", y = Count, fill=Subtype)) +
  geom_col(width = 1, color = NA, alpha = 0.9) +
  coord_polar(theta = "y") +
  scale_fill_observable()+
  geom_text_repel(data = subtype_df2,
                 aes(y = pos, label = label),
                size = 5, nudge_x = 0.85, show.legend = FALSE, fill = NA,
                segment.color = NA)+
  theme_void() +
  theme(legend.position = "none")
subtype_plot
```


```{r}
subtype_df <- data.frame("Stage" = all_data$Stage, "Subtype" = unlist(all_data$Subtype))
subtype_df$Subtype[is.na(subtype_df$Subtype) == T] <- "Not Reported"
```

```{r}
subtype_df2 <- rbind(data.frame(table(subset(subtype_df, Stage == "Stage 0")$Subtype)),
                     data.frame(table(subset(subtype_df, Stage == "Stage I")$Subtype)),
                     data.frame(table(subset(subtype_df, Stage == "Stage II")$Subtype)),
                     data.frame(table(subset(subtype_df, Stage == "Stage III")$Subtype)),
                     data.frame(table(subset(subtype_df, Stage == "Stage IV")$Subtype)),
                     data.frame(table(subset(subtype_df, Stage == "Not Reported")$Subtype)))

subtype_df2$Stage <- c(rep("0", 2), rep("I", 5), rep("II", 5), rep("III", 5), rep("IV", 5),
                       rep("Not Reported", 5))

colnames(subtype_df2) <- c("Subtype", "Count", "Stage")
```

```{r}
pal <- pal_observable("observable10")(10)
```

Figure 1c
```{r}
subtype_df2$Subtype <- factor(subtype_df2$Subtype, levels = c("Not Reported","HR+/HER2-", "HR+/HER2+", "HR-/HER2+", "HR-/HER2-"))
subtype_df2$Stage <- factor(subtype_df2$Stage, levels = c("Not Reported", "0", "I", "II", "III", "IV"))
subtype_df2$Count[subtype_df2$Count < 10] <- NA

subtype_plot <- ggplot(subtype_df2, aes(x = Subtype, y = Count, fill = Stage)) +
  geom_bar(position = "fill", stat = "identity", color = NA, alpha = 0.9) +
  theme_classic() +
  ylab("Proportion of Patients") +
  #scale_fill_manual(values = c("0" = p1_light[1], "I" = p1_light[2], "II" = p1_light[3], "III" = p1_light[4], "IV" = p1_light[5],
   #                                                "Not Reported" = "white")) +
  geom_text(aes(label = Count), size = 6, position = position_fill(vjust = 0.5)) +
  scale_fill_observable()+
  theme(axis.text = element_text(size = 16), axis.title = element_text(size = 16), legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
subtype_plot
```

Upset plots for combinations of patients (Supplementary Figure)
```{r}
tests2 <- subset(all_data, Subtype == "HR-/HER2-")[,c(11:15)]
combo <- function(i, data){
  comb <- paste0(data[i,], collapse = "_")
  return(comb)
}
tests2$Combination <- unlist(lapply(1:nrow(tests2), combo, data = tests2))

combo <- as.data.frame(table(tests2$Combination))
combo <- combo[order(-combo$Freq),]
combo$Var1 <- factor(combo$Var1, levels = combo$Var1)

dat2 <- unique(tests2)

restructure_df <- function(comb, data){
  sub <- subset(data, Combination == comb)
  genes <- c()
  if (sub$Ashion == "Yes"){genes <- c(genes, "Ashion")}
  if (sub$Guardant == "Yes"){genes <- c(genes, "Guardant")}
  if (sub$Tempus == "Yes"){genes <- c(genes, "Tempus")}

  rows <- data.frame("Combination"  = comb, "Test" = genes)
  return(rows)
}

dat3 <- dplyr::bind_rows(lapply(dat2$Combination, restructure_df, data = dat2))
colnames(combo)[1] <- "Combination"
combo <- merge(combo, dat3, by = "Combination")
combo$Test <- factor(combo$Test, levels = rev(c("Ashion", "Guardant", "Tempus")))
combo$Type <- "Somatic Mutations"
```


```{r}
bar <- ggplot(combo, aes(x = Combination, y = Freq)) +
  geom_bar(stat = "identity", position = "dodge", color = NA, fill = pal[1], alpha = 0.9) +
  theme_classic() +
  xlab(NULL) +
  ylab("Number of Patients") +
  ggtitle("HR-/HER2-") +
  theme(axis.text.x = element_blank(), panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.y = element_text(size = 14), axis.title.y = element_text(size = 14), plot.title = element_text(hjust = 0.1, vjust = 0.5, size = 16)) 

point <- ggplot(combo, aes(x = Combination, y = Test)) +
  geom_point(size = 3, color = pal[1], alpha = 0.9) +
  geom_line(aes(group = Combination), color = pal[1], alpha = 0.9) +
  xlab(NULL) +
  theme_minimal()+
  theme(axis.text.x = element_blank(), panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.y = element_text(size = 14), axis.title.y = element_text(size = 14)) 

hrnegher2pos_upset <- ggarrange(bar, point, ncol = 1, nrow = 2, align = "v", heights = c(1,2))
hrnegher2pos_upset
```

Heatmap for treatments before and after genomic test (Figure 1d-e)
```{r}
pcdm_med <- subset(pcdm_med, UUID %in% all_data$UUID)
cr_med <- subset(cr_med, UUID %in% all_data$UUID)
comb_med <- rbind(pcdm_med, cr_med)
comb_tests <- rbind(pcdm_tests, cr_tests)
colnames(comb_tests)[1] <- "UUID"

comb_med2 <- data.frame()
for (pt in unique(comb_tests$`specimenID/accessionID`)){
    sub <- subset(comb_tests, `specimenID/accessionID` == pt & genetic_test %in% c("Ashion", "Guardant", "Tempus"))
    sub_meds <- subset(comb_med, UUID == sub$UUID[1])
    if (nrow(sub_meds) > 0){
    test_day <- sub$days_since_first_dx[1]
    rel_test <- as.numeric(sub_meds$days_Dx_to_Med) - as.numeric(test_day)
    sub_meds$Rel_Test <- rel_test
    sub_meds$Test_ID <- pt
    comb_med2 <- rbind(comb_med2, sub_meds)
    }
}
```

```{r}
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "CDK_inhibitor"] <- "CDK Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_AI"] <- "Hormone Therapy (AI)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_GnRH"] <- "Hormone Therapy (GnRH)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_progestin"] <- "Hormone Therapy (Progestin)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_SERD"] <- "Hormone Therapy (SERD)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "HER2_ERBB_inhibitor"] <- "HER2/ERBB Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_SERM"] <- "Hormone Therapy (SERM)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "MTOR_inhibitor"] <- "mTOR Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Immune_CKI"] <- "Immunotherapy (CKI)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Immune_other"] <- "Immunotherapy (Other)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "VEGF_Inhibitor"] <- "VEGF Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "TK_inhibitor"] <- "TK Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "TROP2_inhibitor"] <- "TROP2 Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "ALK_inhibitor"] <- "ALK Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "PI3K_inhibitor"] <- "PI3K Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "PARP_inhibitor"] <- "PARP Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "MEK_inhibitor"] <- "MEK Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Hormone_AR"] <- "Hormone Therapy (AR)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Proteosome_Inhibitor"] <- "Proteosome Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "EGFR_inhibitor"] <- "EGFR Inhibitor"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "HER2_ERBB_Inhibitor (ADC)"] <- "HER2/ERBB Inhibitor (ADC)"

comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Microtubule_Inhibitor"] <- "Chemotherapy (Microtubule Inhibitor)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Alkylating_Agent"] <- "Chemotherapy (Alkylating Agent)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Platinum_Based_Chemotherapy"] <- "Chemotherapy (Platinum Based)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Topoisomerase_Inhibitor"] <- "Chemotherapy (Topoisomerase Inhibitor)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Chemotherapy_other"] <- "Chemotherapy (Other)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Monoclonal_Antibody"] <- "Monoclonal Antibody"

comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Antimetabolite"] <- "Chemotherapy (Antimetabolite)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Taxane"] <- "Chemotherapy (Taxane)"
comb_med2$Drug_Classification_V4[comb_med2$Drug_Classification_V4 == "Antracycline"] <- "Chemotherapy (Antracycline)"

drug_list <- na.omit(unique(comb_med2$Drug_Classification_V4))
```


Medications Before (Figure 1d)
```{r}
get_before_meds <- function(id, data){
  sub <- subset(data, Test_ID == id)
  drugs <- c()
  for (drug in unique(sub$Drug_Classification_V4)){
    sub2 <- subset(sub, Drug_Classification_V4 == drug)
    days <- na.omit(sub2$Rel_Test)
    if (length(days) > 0){
    min <- min(days)
    if (min < -90){drugs <- c(drugs, drug)}
    }
  }
  if (length(drugs) > 0){
  row <- data.frame("ID" = id, "Drug_Classification" = drugs, "UUID" = sub$UUID[1])
  return(row)
  }
}

before <- dplyr::bind_rows(lapply(unique(comb_med2$Test_ID), get_before_meds, data = comb_med2))
```

```{r}
reformat_data <- function(id, data, drugs){
  sub <- subset(data, ID == id)
  val <- c()
  for (d in drugs){
    if (d %in% sub$Drug_Classification){val <- c(val, d)}
    else{val <- c(val, NA)}
  }
  row <- data.frame(matrix(nrow = 1, ncol = length(drugs)+2))
  row[1,] <- c(id, sub$UUID[1], val)
  colnames(row) <- c("ID", "UUID", drugs)
  return(row)
}

before2 <- dplyr::bind_rows(lapply(before$ID, reformat_data, data = before, drugs = drug_list))
before2 <- unique(before2)

all_sub <- subset(all_data, UUID %in% before2$UUID)
before2 <- merge(all_sub[,c(1,3,9)], before2, by = "UUID")
before2$Subtype[is.na(before2$Subtype) == T] <- "Not Reported"
```

```{r}
# Calculate the proportion of NA values in each column
na_proportions <- colMeans(is.na(before2))

# Identify columns with more than 95% NA values
cols_to_remove <- na_proportions > 0.99

# Remove identified columns from the data frame
before2_plot <- before2[, !cols_to_remove]
```

```{r}
pal <- pal_observable("observable10", alpha = 0.9)(10)
```


```{r}
library(ComplexHeatmap)

drug_list <- setdiff(colnames(before2_plot)[-c(1:4)], c("NSAID", "Other"))
before2_plot <- before2_plot[,c(colnames(before2_plot)[1:4], drug_list)]
drug_list <- drug_list[order(drug_list)]
hm_col = pal[8]
colors = c("CDK Inhibitor" = hm_col, "Chemotherapy (Alkylating Agent)" = hm_col,
           "Chemotherapy (Antimetabolite)" = hm_col, "Chemotherapy (Antracycline)" = hm_col,
           "Chemotherapy (Microtubule Inhibitor)" = hm_col, "Chemotherapy (Platinum Based)" = hm_col,
           "Chemotherapy (Taxane)" = hm_col, "Chemotherapy (Topoisomerase Inhibitor)" = hm_col,
           "HER2/ERBB Inhibitor" = hm_col, "HER2/ERBB Inhibitor (ADC)" = hm_col,
           "Hormone Therapy (AI)" = hm_col, "Hormone Therapy (GnRH)" = hm_col,
           "Hormone Therapy (SERD)" = hm_col, "Hormone Therapy (SERM)" = hm_col,
           "Immunotherapy (CKI)" = hm_col, "mTOR Inhibitor" = hm_col,
           "PARP Inhibitor" = hm_col, "PI3K Inhibitor" = hm_col,
           "TROP2 Inhibitor" = hm_col)
before2_plot$Subtype <- factor(before2_plot$Subtype, levels = c("HR+/HER2-", "HR+/HER2+", "HR-/HER2+", "HR-/HER2-", "Not Reported"))

jpeg("meds_before.jpeg", width = 2500, height = 1500, units = "px", res = 300) # You can adjust width and height

column_ha = HeatmapAnnotation(Stage = before2_plot$Stage, Subtype = before2_plot$Subtype,
                              col = list(Stage = c("Stage 0" = pal[5],
                                                   "Stage I" = pal[6],
                                                   "Stage II" = pal[7],
                                                   "Stage III" = pal[8],
                                                   "Stage IV" = pal[9],
                                                   "Not Reported" = "white"),
                                         Subtype = c("HR+/HER2-" = pal[4],
                                                   "HR+/HER2+" = pal[3],
                                                    "HR-/HER2+" = pal[2],
                                                   "HR-/HER2-" = pal[1],
                                                   "Not Reported" = "white")))
Heatmap(t(before2_plot[,5:ncol(before2_plot)]), name = "Medications_Before", col = colors, na_col = "white",  top_annotation = column_ha, show_heatmap_legend = FALSE, show_column_names = FALSE, column_split = before2_plot$Subtype, row_order = drug_list, column_title_rot = 45,
                      cluster_column_slices = T, cluster_columns = T, use_raster = F, row_names_gp = gpar(fontsize = 8))
dev.off()
```


Medications (Any) After (Figure 1e)
```{r}
get_after_meds <- function(id, data){
  sub <- subset(data, Test_ID == id)
  drugs <- c()
  for (drug in unique(sub$Drug_Classification_V4)){
    sub2 <- subset(sub, Drug_Classification_V4 == drug)
    days <- na.omit(sub2$Rel_Test)
    if (length(days) > 0){
    max <- max(days)
    min <- min(days)
    if (min >= -90){drugs <- c(drugs, drug)}
    }
  }
  if (length(drugs) > 0){
  row <- data.frame("ID" = id, "Drug_Classification" = drugs, "UUID" = sub2$UUID[1])
  return(row)
  }
}

after_all <- dplyr::bind_rows(lapply(unique(comb_med2$Test_ID), get_after_meds, data = comb_med2))
```

```{r}
reformat_data <- function(id, data, drugs){
  sub <- subset(data, ID == id)
  val <- c()
  for (d in drugs){
    if (d %in% sub$Drug_Classification){val <- c(val, d)}
    else{val <- c(val, NA)}
  }
  row <- data.frame(matrix(nrow = 1, ncol = length(drugs)+2))
  row[1,] <- c(id, sub$UUID[1], val)
  colnames(row) <- c("ID", "UUID", drugs)
  return(row)
}

after2_all <- dplyr::bind_rows(lapply(after_all$ID, reformat_data, data = after_all, drugs = drug_list))
after2_all <- unique(after2_all)

all_sub <- subset(all_data, UUID %in% after2_all$UUID)
after2_all <- merge(all_sub[,c(1,3,9)], after2_all, by = "UUID")
after2_all$Subtype[is.na(after2_all$Subtype) == T] <- "Not Reported"

```

```{r}
hm_col = pal[6]
colors = c("CDK Inhibitor" = hm_col, "Chemotherapy (Alkylating Agent)" = hm_col,
           "Chemotherapy (Antimetabolite)" = hm_col, "Chemotherapy (Antracycline)" = hm_col,
           "Chemotherapy (Microtubule Inhibitor)" = hm_col, "Chemotherapy (Platinum Based)" = hm_col,
           "Chemotherapy (Taxane)" = hm_col, "Chemotherapy (Topoisomerase Inhibitor)" = hm_col,
           "HER2/ERBB Inhibitor" = hm_col, "HER2/ERBB Inhibitor (ADC)" = hm_col,
           "Hormone Therapy (AI)" = hm_col, "Hormone Therapy (GnRH)" = hm_col,
           "Hormone Therapy (SERD)" = hm_col, "Hormone Therapy (SERM)" = hm_col,
           "Immunotherapy (CKI)" = hm_col, "mTOR Inhibitor" = hm_col,
           "PARP Inhibitor" = hm_col, "PI3K Inhibitor" = hm_col,
           "TROP2 Inhibitor" = hm_col)
after2_all_plot <- after2_all[,colnames(before2_plot)]
after2_all_plot$Subtype <- factor(after2_all_plot$Subtype, levels = c("HR+/HER2-", "HR+/HER2+", "HR-/HER2+", "HR-/HER2-", "Not Reported"))

jpeg("meds_after.jpeg", width = 2500, height = 1500, units = "px",
     res = 300) # You can adjust width and height

column_ha = HeatmapAnnotation(Stage = after2_all_plot$Stage, Subtype = after2_all_plot$Subtype,
                              col = list(Stage = c("Stage 0" = pal[5],
                                                   "Stage I" = pal[6],
                                                   "Stage II" = pal[7],
                                                   "Stage III" = pal[8],
                                                   "Stage IV" = pal[9],
                                                   "Not Reported" = "white"),
                                         Subtype = c("HR+/HER2-" = pal[4],
                                                   "HR+/HER2+" = pal[3],
                                                    "HR-/HER2+" = pal[2],
                                                   "HR-/HER2-" = pal[1],
                                                   "Not Reported" = "white")))
Heatmap(t(after2_all_plot[,5:ncol(after2_all_plot)]), name = "Medications_After", col = colors, na_col = "white",  top_annotation = column_ha, show_heatmap_legend = FALSE, show_column_names = FALSE, column_split = after2_all_plot$Subtype,column_title_rot = 45,
                     row_order = drug_list, use_raster = F, row_names_gp = gpar(fontsize = 8))
dev.off()
```

```{r}
before3 <- before2[,c(1,4:ncol(before2))]
colnames(before3)[-c(1,2)] <- paste0(colnames(before3)[-c(1:2)], "_Before")
cr_tests2 <- subset(cr_tests2, genetic_test %in% c("Guardant", "Tempus", "ashion"))
pcdm_tests2 <- subset(pcdm_tests2, source %in% c("Guardant", "Tempus", "ashion"))

no_before <- setdiff(setdiff(c(cr_tests2$`specimenID/accessionID`, pcdm_tests2$`specimenID/accessionID`), before3$ID), "NULL")
missing_dat <- data.frame(matrix(NA, ncol = ncol(before3), nrow = length(no_before)))
colnames(missing_dat) <- colnames(before3)
missing_tests <- unique(rbind(subset(cr_tests2, `specimenID/accessionID` %in% no_before)[,c(1,3)], subset(pcdm_tests2, `specimenID/accessionID` %in% no_before)[,c(1,3)]))
missing_tests <- subset(missing_tests, `specimenID/accessionID` != "NULL")
missing_dat$UUID <- missing_tests$uuid
missing_dat$ID <- missing_tests$`specimenID/accessionID`
before3 <- rbind(before3, missing_dat)

after3 <- after2_all[,c(1,4:ncol(after2_all))]
colnames(after3)[-c(1,2)] <- paste0(colnames(after3)[-c(1:2)], "_After")

no_after <- setdiff(setdiff(c(cr_tests2$`specimenID/accessionID`, pcdm_tests2$`specimenID/accessionID`), after3$ID), "NULL")
missing_dat <- data.frame(matrix(NA, ncol = ncol(after3), nrow = length(no_after)))
colnames(missing_dat) <- colnames(after3)
missing_tests <- unique(rbind(subset(cr_tests2, `specimenID/accessionID` %in% no_after)[,c(1,3)], subset(pcdm_tests2, `specimenID/accessionID` %in% no_after)[,c(1,3)]))
missing_tests <- subset(missing_tests, `specimenID/accessionID` != "NULL")
missing_dat$UUID <- missing_tests$uuid
missing_dat$ID <- missing_tests$`specimenID/accessionID`
after3 <- rbind(after3, missing_dat)

treatment3 <- unique(merge(before3, after3[,-1], by = "ID"))


all_data2 <- merge(all_data, treatment3, by = "UUID")
```

Adding time on treatment for pretreatments
```{r}
get_before_med_times <- function(id, data){
  sub <- subset(data, Test_ID == id)
  drugs <- c()
  total_days <- c()
  for (drug in unique(sub$Drug_Classification_V4)){
    sub2 <- subset(sub, Drug_Classification_V4 == drug)
    days <- na.omit(sub2$Rel_Test)
    if (length(days) > 0){
    min <- min(days)
    if (min < -90){
      drugs <- c(drugs, drug)
      time_start <- min
      time_end <- c(-90, max(days))[which.min(c(-90, max(days)))]
      total_days <- c(total_days, length(time_start:time_end))
      }
    }
  }
  if (length(drugs) > 0){
  row <- data.frame("ID" = id, "Drug_Classification" = drugs, "UUID" = sub$UUID[1], "Time_on_Med" = total_days)
  return(row)
  }
}

before_times <- dplyr::bind_rows(lapply(unique(comb_med2$Test_ID), get_before_med_times, data = comb_med2))
```

```{r}
reformat_data <- function(id, data, drugs){
  sub <- subset(data, ID == id)
  val <- c()
  for (d in drugs){
    if (d %in% sub$Drug_Classification){
      sub2 <- subset(sub, Drug_Classification == d)
      val <- c(val, sub2$Time_on_Med)
      }
    else{val <- c(val, 0)}
  }
  row <- data.frame(matrix(nrow = 1, ncol = length(drugs)+2))
  row[1,] <- c(id, sub$UUID[1], val)
  colnames(row) <- c("ID", "UUID", drugs)
  return(row)
}

before2_times <- dplyr::bind_rows(lapply(before_times$ID, reformat_data, data = before_times, drugs = drug_list))
before2_times <- unique(before2_times)
colnames(before2_times)[-c(1:2)] <- paste0("Duration of ", colnames(before2_times)[-c(1:2)])

all_sub <- subset(all_data, UUID %in% before2_times$UUID)
before2_times <- merge(all_sub[,c(1,3,9)], before2_times, by = "UUID")
before2_times$Subtype[is.na(before2_times$Subtype) == T] <- "Not Reported"

#write.csv(before2_times, "pretreatment_durations.csv", row.names = F)
```

Save all curated clinical data
```{r}
#write.csv(all_data2, "combined_clinical.csv", row.names = F)
```


Calculate time to treatment change
```{r}
get_after_meds <- function(id, data){
  sub <- subset(data, Test_ID == id)
  drugs <- c()
  start_day <- c()
  med_names <- c()
  for (drug in unique(sub$Drug_Classification_V4)){
    sub2 <- subset(sub, Drug_Classification_V4 == drug)
    days <- na.omit(sub2$Rel_Test)
    
    if (length(days) > 0){
    max <- max(days)
    min <- min(days)
    if (max > -90){
      drugs <- c(drugs, drug)
      med_names <- c(med_names, paste0(unique(sub2$med_name), collapse = "_"))
      start_day <- c(start_day, min(days))
      }
    }
  }
  if (length(drugs) > 0){
  row <- data.frame("ID" = id, "Drug_Classification" = drugs, 
                    "Start_Time" = start_day, "Med_Name" = med_names, "UUID" = sub2$UUID[1])
  return(row)
  }
}

after_all <- dplyr::bind_rows(lapply(unique(comb_med2$Test_ID), get_after_meds, data = comb_med2))
```


```{r}
pcdm_cohort <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "PCDM_Cohort")
cr_cohort <- read_excel("~//Documents//COH Breast Cohort//HB-1646_BreastCaCohort_v5.xlsx", sheet = "CR_Cohort")

colnames(pcdm_cohort) <- pcdm_cohort[1,]
pcdm_cohort <- pcdm_cohort[-1,]

colnames(cr_cohort) <- cr_cohort[1,]
cr_cohort <- cr_cohort[-1,]

comb_last <- rbind(cr_cohort[,c(1,7)], pcdm_cohort[,c(1,6)])
```

```{r}
sixm_combos <- function(x, data, test_data, last_day, treat_data){
  # subset test data by individual test
  test <- subset(test_data, `specimenID/accessionID` == x & genetic_test %in% c("ashion", "Tempus", "Guardant"))
  uuid <- test$uuid[1]
  test_type <- test$genetic_test[1]
  # subset overall survival and treatment data from uuid
  last_day <- subset(last_day, UUID == uuid)
  last_treat <- subset(treat_data, UUID == uuid)
  # find last day known alive
  last <- max(c(as.numeric(last_day$days_Dx_to_LastAppt), as.numeric(last_treat$days_Dx_to_Med)))
  if (nrow(test) > 0){
    test <- test[order(test$days_since_first_dx),]
    test_time <- test$days_since_first_dx[1]
  
  # get all medications taken at time of sequencing window or after
  sub <- subset(data, ID == x)
  sub <- subset(sub, Start_Time >= -90)
  if (nrow(sub) > 0){
  start_time <- min(as.numeric(sub$Start_Time))
  end_time <- 90
  # get first combination of treatments
  firstcomb <- subset(sub, Start_Time %in% start_time:end_time)
  start_times <- firstcomb$Start_Time
  firstcomb_medname <- unique(firstcomb$Med_Name)
  firstcomb <- unique(firstcomb$Drug_Classification)
  
  
  sub <- subset(sub, Drug_Classification %in% c("HER2/ERBB Inhibitor", "CDK Inhibitor", "Immunotherapy (CKI)", "Chemotherapy (Microtubule Inhibitor)", "Chemotherapy (Alkylating Agent)", "Chemotherapy (Taxane)", "Chemotherapy (Antimetabolite)",
                                                "Chemotherapy (Topoisomerase Inhibitor)", "Chemotherapy (Antracycline)",
                                                "VEGF Inhibitor", "TK Inhibitor", "mTOR Inhibitor", "PI3K Inhibitor",
                                                "TROP2 Inhibitor", "PARP Inhibitor", "ALK Inhibitor", "Immunotherapy (Other)",
                                                "EGFR Inhibitor", "MEK Inhibitor", "HER2/ERBB Inhibitor (ADC)"))
  
  # find any medication started for the first time after sequencing window
  later_meds <- subset(sub, Start_Time >= end_time)
  later_meds.rm <- subset(later_meds, Drug_Classification %in% firstcomb)
  later_meds <- dplyr::setdiff(later_meds, later_meds.rm)

  # if there are medications in list, medications were changes event = 1, time is from first start date of new meds
  if (nrow(later_meds) > 0){
  time <- min(later_meds$Start_Time) - min(start_times)
  event <- 1
  }

  # if list is empty, event = 0 no indicated treatment change, time is overall follow up time relative to start date of medication
  if (nrow(later_meds) == 0){
  time <- last - as.numeric(test_time)
  time <- time - min(start_times)
  event <- 0
  }
  
  comb <- firstcomb
  row <- data.frame("ID" = x, "UUID" = uuid,  "Medication_Combination" = comb, "Event" = event, "Time" = time, "Med_Name" = firstcomb_medname)
  return(row)
  }
  }
}
colnames(pcdm_tests2)[2] <- "genetic_test"

after_combinations <- dplyr::bind_rows(lapply(unique(after_all$ID), sixm_combos, data = after_all, test_data = rbind(cr_tests2, pcdm_tests2),
                                       last_day = comb_last, treat_data = rbind(cr_med, pcdm_med)))
```

```{r}
#write.csv(after_combinations, "patient_treatment_responses_time_to_change.csv", row.names = F)
```

